---
title: "Getting started with genhaz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with genhaz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

## Introduction

In survival analysis, covariate effects are typically modelled through one of
two mechanisms: a multiplicative effect on the hazard (proportional hazards,
PH) or a multiplicative effect on time (accelerated failure time, AFT).
**Generalized hazard (GH) models** combine both simultaneously:

$$h(t \mid x) = h_0\!\left(t\,e^{\beta_1^\top x}\right) \cdot e^{(\beta_1+\beta_2)^\top x}$$

This nests PH ($\beta_1 = 0$), AFT ($\beta_2 = 0$), and additive hazards
(AH, $\beta_1 = -\beta_2$) as special cases. A key advantage: combining the
time-acceleration parameter $\beta_1$ and the hazard-scaling parameter
$\beta_2$ can capture time-varying hazard ratios with only two parameters per
covariate.

`genhaz` fits penalised cubic restricted spline GH models on the log-time
scale, with the smoothing parameter selected automatically by minimising a
modified likelihood cross-validation (LCV) criterion.

## Model types

The `model_type` argument selects the sub-model per covariate:

| `model_type` | Constraint | Interpretation |
|---|---|---|
| `"GH"` | none | Full generalised hazard |
| `"PH"` | $\beta_1 = 0$ | Proportional hazards |
| `"AFT"` | $\beta_2 = 0$ | Accelerated failure time |
| `"AH"` | $\beta_1 = -\beta_2$ | Additive hazards |

Mixed models are supported via a vector, e.g. `model_type = c("PH", "GH")`
for two covariates.

## Censoring types

| `Surv` call | Censoring | Internal type |
|---|---|---|
| `Surv(time, event)` | Right-censoring | `"rc"` |
| `Surv(start, stop, event)` | Left-truncation + right-censoring | `"lt_rc"` |
| `Surv(t1, t2, type="interval2")` | Interval censoring | `"ic"` |

---

Note: only rc is tested at this point in time.

## Real-data example: melanoma survival

We use the `biostat3::melanoma` dataset: 7,775 patients with melanoma cancer,
including age group, period of diagnosis, sex, cancer stage, survival time in
months, and a death indicator. The question of interest is the effect of
non-localised stage on survival, adjusted for age, period, and sex.

### Data preparation

```{r melanoma-data, message=FALSE, eval=FALSE}
library(biostat3)

mel        <- biostat3::melanoma
mel$X      <- ifelse(mel$stage == "Localised", 0, 1)
mel$event  <- ifelse(mel$status == "Dead: cancer", 1, 0)
mel$time   <- mel$surv_mm
mel$period <- ifelse(mel$year8594 == "Diagnosed 75-84", 0, 1)
```

### Fitting the model

The full model fit takes approximately 9 minutes and is not re-run here.
The code below shows exactly what was run to produce the stored result:

```{r melanoma-fit, eval=FALSE}
fit_melanoma <- fit_genhaz(
  Surv(mel$time, mel$event), ~ X + period + agegrp + sex,
  data       = mel,
  model_type = "GH",
  profile    = TRUE,
  n_knots    = 8,
  tol_LCV    = 0.001,
  lcv_method = "optimize"
)
```

```{r melanoma-load, message=FALSE}
library(genhaz)
library(survival)
data("fit_melanoma")
```

```{r melanoma-setup, message=FALSE}
library(biostat3)
mel        <- biostat3::melanoma
mel$X      <- ifelse(mel$stage == "Localised", 0, 1)
mel$period <- ifelse(mel$year8594 == "Diagnosed 75-84", 0, 1)
new_time   <- seq(0.5, 320, by = 0.5)   # avoids t = 0 (needed for time_ratio)

nd_mel <- data.frame(
  X      = c(1L, 0L),
  period = c(1L, 1L),
  agegrp = factor(c("60-74", "60-74"), levels = levels(mel$agegrp)),
  sex    = factor(c("Male",  "Male"),  levels = levels(mel$sex))
)
rownames(nd_mel) <- c("Non-Localised", "Localised")
```

### Results

```{r melanoma-print}
print(fit_melanoma)
```

```{r melanoma-summary}
summary(fit_melanoma)
```

Patients with non-localised disease progress through the baseline hazard
$\exp(\hat\beta_1)$ times faster and face a $\exp(\hat\beta_2)$ times higher
hazard at every time point. Wald CIs for each:

```{r melanoma-waldci}
confint(fit_melanoma, "beta1_X")
confint(fit_melanoma, "beta2_X")
```

### Hazard and survival curves

`plot()` dispatches on the fitted model, calls `predict()` internally, and
sets `ylim` from the full CI range so confidence bands are never clipped.
Evaluated at age group 60–74, male sex, diagnosed 1985–94.

```{r melanoma-hazard-plot}
plot(fit_melanoma, newdata = nd_mel, times = new_time, type = "hazard",
     interval = "confidence", col = c("steelblue", "firebrick"),
     xlab = "Time (months)", main = "Estimated hazard — melanoma, GH model")
```

```{r melanoma-survival-plot}
plot(fit_melanoma, newdata = nd_mel, times = new_time, type = "survival",
     interval = "confidence", col = c("steelblue", "firebrick"),
     xlab = "Time (months)", main = "Estimated survival — melanoma, GH model")
```

### Time-varying hazard ratio

`type = "hazard_ratio"` gives $h_1(t)/h_0(t)$ (exposed over baseline) with a
log-scale delta-method CI. Under PH this would be flat; the GH model captures
the time variation.
`plot()` works directly on the `predict()` result.

```{r melanoma-hr}
hr_mel <- predict(fit_melanoma, newdata = nd_mel,
                  times = seq(0.5, 200, by = 0.5), type = "hazard_ratio",
                  interval = "confidence")
plot(hr_mel, col = "purple",
     xlab = "Time (months)",
     main = "Time-varying HR — non-localised vs localised")
abline(h = 1, lty = 2, col = "grey50")
```

With only two parameters for the stage effect, the GH model captures the
time-varying hazard ratio: high at diagnosis, levelling off after approximately
6 years.

### Survival difference and RMST

`type = "surv_diff"` gives $S_1(t) - S_0(t)$ (exposed minus baseline) on the
linear scale; the delta-method accounts for the correlation between the two
curves.
`type = "rmst_diff"` integrates this difference up to each restriction
time $\tau$ via 25-point Gauss-Legendre quadrature.

```{r melanoma-differences}
diff_s_mel <- predict(fit_melanoma, newdata = nd_mel,
                      times = new_time, type = "surv_diff",
                      interval = "confidence")
plot(diff_s_mel,
     xlab = "Time (months)",
     main = "S(t | Non-localised) − S(t | Localised)")
abline(h = 0, lty = 3, col = "grey70")

rmst_mel <- predict(fit_melanoma, newdata = nd_mel,
                    times = new_time, type = "rmst_diff",
                    interval = "confidence")
plot(rmst_mel,
     xlab = "Time (months)",
     main = "RMST(t | Non-localised) − RMST(t | Localised)")
abline(h = 0, lty = 3, col = "grey70")
```

### Time ratio

`type = "time_ratio"` gives $\tau/t$ where $S_0(\tau) = S_1(t)$ (the relation
$S(t\mid x_1)=S_0(\tau(t))$): the baseline time $\tau$ at which the localised
(baseline, group 0, row 2) group reaches the survival level the non-localised
(exposed, group 1, row 1) group has at time $t$. The non-localised group fares
worse, so the localised group needs *longer* to fall that far, giving
$\tau > t$. Each $\tau$ is found via `uniroot`; the delta-method CI uses the
implicit function theorem. The search is capped at the model support
`exp(max(fit$knots))` (`tau_max`); the localised baseline survival plateaus
above the non-localised level, so beyond an early window no $\tau$ exists
(returned `NA`) — we therefore use an in-support grid.

```{r melanoma-time-ratio}
tr_mel <- predict(fit_melanoma, newdata = nd_mel,
                  times = seq(2, 24, by = 2), type = "time_ratio",
                  interval = "confidence")
plot(tr_mel, col = "darkorange",
     xlab = "Time (months)", main = "Time ratio — non-localised vs localised")
abline(h = 1, lty = 2, col = "grey50")
```

### Acceleration factor

`type = "acc_factor"` reports the time-varying acceleration **effect** on the
log scale,
$$f(t) = \log \tau'(t) = \log h_1(t) - \log h_0(\tau(t)),$$
the time-varying analogue of $\beta_1$: it is defined so that
$$S(t \mid x_1) = S_0\!\left(\int_0^t e^{f(u)\,X}\,du\right),$$
i.e. the warped time is $\tau(t)=\int_0^t e^{f(u)X}du$ with instantaneous rate
$e^{f(t)X}=\tau'(t)$, and $f(t)=\beta_1$ in the constant-AFT limit. So $f(t)$ is
directly comparable to the fitted $\hat\beta_1$ (here $\approx 1.14$). The
delta-method CI uses the implicit function theorem together with the hazard
time-derivative `post(..., "dh_dt")`; the same support cap applies, so we use
the in-support grid.

This is a work in progress.

```{r melanoma-acc-factor}
af_mel <- predict(fit_melanoma, newdata = nd_mel,
                  times = seq(2, 24, by = 2), type = "acc_factor",
                  interval = "confidence")
plot(af_mel, col = "purple",
     xlab = "Time (months)", main = "Acceleration effect f(t) — non-localised vs localised")
abline(h = fit_melanoma$par["beta1_X"], lty = 2, col = "grey50")  # constant-AFT beta1
```


## Simulation and benchmarking

`sim_scenario()` generates right-censored survival data from one of three
named scenarios with mixture-Weibull baseline hazards (bathtub, hump-shaped,
early peak). `mixWeibSc()` evaluates the corresponding true $h$, $H$, and $S$
on a grid for comparison with fitted estimates.

```{r simulate}
set.seed(82359)
dat <- sim_scenario(scenario = 1, beta1 = 0.5, beta2 = 0.5, n = 2000)
head(dat)
```

```{r fit}
fit <- fit_genhaz(
  surv       = Surv(dat$time, dat$event),
  formula    = ~ X,
  data       = dat,
  model_type = "GH",
  profile    = TRUE,
  n_knots    = 6,
  tol_LCV    = 0.05,
  lcv_method = "optimize"
)
```

```{r print}
print(fit)
```

### Inference

`confint()` gives Wald CIs for the parameters; with `diff = TRUE` it returns a
CI for $\beta_1 - \beta_2$ (accounting for their covariance).

```{r waldci}
confint(fit, "beta1_X")
confint(fit, "beta2_X")
confint(fit, c("beta1_X", "beta2_X"), diff = TRUE)
```

Fit a restricted (PH) model and test against the full GH model:

```{r lr-test}
fit_ph <- fit_genhaz(
  surv       = Surv(dat$time, dat$event),
  formula    = ~ X,
  data       = dat,
  model_type = "PH",
  profile    = TRUE,
  n_knots    = 6,
  tol_LCV    = 0.05,
  lcv_method = "optimize"
)

anova(fit_ph, fit)
```

---

## Smoothing-parameter selection

The smoothing parameter $\lambda$ is chosen by minimising the modified LCV
criterion. Three strategies are available via `lcv_method`:

| `lcv_method` | Method | Notes |
|---|---|---|
| `"full"` (default) | Root-find on full LCV gradient (third-derivative correction) | Most accurate |
| `"approx"` | Root-find on first-order LCV gradient | Faster |
| `"optimize"` | Direct `optimize()` on LCV, no gradient | Gradient-free |

```{r lcv-methods, eval=FALSE}
# First-order gradient (faster)
fit <- fit_genhaz(..., profile = TRUE, lcv_method = "approx")

# Pure optimisation (no gradient)
fit <- fit_genhaz(..., profile = TRUE, lcv_method = "optimize")
```
