---
title: "Getting started with genhaz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with genhaz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

## Introduction

In survival analysis, covariate effects are typically modelled through one of
two mechanisms: a multiplicative effect on the hazard (proportional hazards,
PH) or a multiplicative effect on time (accelerated failure time, AFT).
**Generalized hazard (GH) models** combine both simultaneously:

$$h(t \mid x) = h_0\!\left(t\,e^{\beta_1^\top x}\right) \cdot e^{(\beta_1+\beta_2)^\top x}$$

This nests PH ($\beta_1 = 0$), AFT ($\beta_2 = 0$), and additive hazards
(AH, $\beta_1 = -\beta_2$) as special cases. A key advantage: combining the
time-acceleration parameter $\beta_1$ and the hazard-scaling parameter
$\beta_2$ can capture time-varying hazard ratios with only two parameters per
covariate.

`genhaz` fits penalised cubic restricted spline GH models on the log-time
scale, with the smoothing parameter selected automatically by minimising a
modified likelihood cross-validation (LCV) criterion.

## Model types

The `model_type` argument selects the sub-model per covariate:

| `model_type` | Constraint | Interpretation |
|---|---|---|
| `"GH"` | none | Full generalised hazard |
| `"PH"` | $\beta_1 = 0$ | Proportional hazards |
| `"AFT"` | $\beta_2 = 0$ | Accelerated failure time |
| `"AH"` | $\beta_1 = -\beta_2$ | Additive hazards |

Mixed models are supported via a vector, e.g. `model_type = c("PH", "GH")`
for two covariates.

## Censoring types

| `Surv` call | Censoring | Internal type |
|---|---|---|
| `Surv(time, event)` | Right-censoring | `"rc"` |
| `Surv(start, stop, event)` | Left-truncation + right-censoring | `"lt_rc"` |
| `Surv(t1, t2, type="interval2")` | Interval censoring | `"ic"` |

---

## Quick start

We simulate 500 observations from scenario 1 (bathtub-shaped baseline hazard)
with $\beta_1 = \beta_2 = 0.5$ and fit a GH model with automatic smoothing
parameter selection.

```{r simulate}
library(genhaz)
library(survival)

set.seed(42)
dat <- sim_scenario(scenario = 1, beta1 = 0.5, beta2 = 0.5, n = 500)
head(dat)
```

```{r fit}
fit <- fit_genhaz(
  surv       = Surv(dat$time, dat$event),
  formula    = ~ X,
  data       = dat,
  model_type = "GH",
  profile    = TRUE,
  n_knots    = 6,
  tol_LCV    = 0.05,
  lcv_method = "optimize"
)
```

---

## Inspecting the fit

`print()` gives a concise overview: model metadata and a coefficient table
with Wald confidence intervals.

```{r print}
print(fit)
```

`summary()` adds the stored formula, exponentiated estimates, and a full CI
table. `exp(beta1)` is the time-acceleration factor; `exp(beta2)` is the
hazard-scaling factor.

```{r summary}
summary(fit)
```

The summary object can also be used programmatically:

```{r summary-access}
s <- summary(fit)
s$coef_tab
```

---

## Predictions

`predict()` accepts a `data.frame` of covariate values (one row per group)
and a time grid, and returns a long `data.frame` with pointwise delta-method
confidence intervals. Row names of `newdata` become the group labels.

```{r predict}
t_grid <- seq(0.01, 8, length.out = 200)

nd <- data.frame(X = c(0, 1))
rownames(nd) <- c("X = 0", "X = 1")

pred_h <- predict(fit, newdata = nd, times = t_grid, type = "hazard")
head(pred_h)
```

Survival and cumulative hazard are requested via `type`:

```{r predict-survival}
pred_s <- predict(fit, newdata = nd, times = t_grid, type = "survival")
head(pred_s)
```

---

## Plotting

`plot()` draws all groups as coloured lines with dashed confidence bands.
Row names of `newdata` appear in the legend.

```{r plot-hazard}
plot(fit, newdata = nd, times = t_grid)
```

```{r plot-survival}
plot(fit, newdata = nd, times = t_grid, type = "survival",
     col = c("steelblue", "firebrick"))
```

Pass additional graphical parameters directly through `...`:

```{r plot-custom}
plot(fit, newdata = nd, times = t_grid, type = "survival",
     col  = c("steelblue", "firebrick"),
     main = "Estimated survival — GH model, scenario 1",
     xlab = "Time (years)",
     ci   = FALSE)
```

---

## Inference

### Wald confidence intervals

`waldCI()` computes a Wald CI for a single parameter; `waldCI_minus()`
provides a CI for the difference $\beta_1 - \beta_2$ (accounting for their
covariance via the delta method).

```{r waldci}
waldCI(fit, "beta1_X")
waldCI(fit, "beta2_X")
waldCI_minus(fit, "beta1_X", "beta2_X")
```

### Likelihood ratio test

Fit a restricted (PH) model and test against the full GH model:

```{r lr-test}
fit_ph <- fit_genhaz(
  surv       = Surv(dat$time, dat$event),
  formula    = ~ X,
  data       = dat,
  model_type = "PH",
  profile    = TRUE,
  n_knots    = 6,
  tol_LCV    = 0.05,
  lcv_method = "optimize"
)

LR(fit_ph, fit)
```

---

## Real-data example: melanoma survival

We use the `biostat3::melanoma` dataset: 7,775 patients with melanoma cancer,
including age group, period of diagnosis, sex, cancer stage, survival time in
months, and a death indicator. The question of interest is the effect of
non-localised stage on survival, adjusted for age, period, and sex.

### Data preparation

```{r melanoma-data, message=FALSE, eval=FALSE}
library(biostat3)

mel        <- biostat3::melanoma
mel$X      <- ifelse(mel$stage == "Localised", 0, 1)
mel$event  <- ifelse(mel$status == "Dead: cancer", 1, 0)
mel$time   <- mel$surv_mm
mel$period <- ifelse(mel$year8594 == "Diagnosed 75-84", 0, 1)
```

### Fitting the model

The full model fit takes approximately ~9 min and is not re-run here.
The code below shows exactly what was run to produce the stored result:

```{r melanoma-fit, eval=FALSE}
fit_melanoma <- fit_genhaz(
  Surv(mel$time, mel$event), ~ X + period + agegrp + sex,
  data       = mel,
  model_type = "GH",
  profile    = TRUE,
  n_knots    = 8,
  tol_LCV    = 0.001,
  lcv_method = "optimize"
)
```

```{r melanoma-load}
data("fit_melanoma")
```

### Results

```{r melanoma-print}
print(fit_melanoma)
```

```{r melanoma-summary}
summary(fit_melanoma)
```

Patients with non-localised disease progress through the baseline hazard
$\exp(\hat\beta_1)$ times faster and face a $\exp(\hat\beta_2)$ times higher
hazard at every time point. Wald CIs for each:

```{r melanoma-waldci}
waldCI(fit_melanoma, "beta1_X")
waldCI(fit_melanoma, "beta2_X")
```

### Hazard and survival curves

We evaluate at age group 60–74, male sex, diagnosed 1985–94 (the adjusted
covariate vector in design-matrix order: X, period, agegrp60-74, sex Male).

```{r melanoma-ci}
new_time <- seq(0, 320, by = 0.5)

CIs_loc     <- CI(fit_melanoma, new_time, c(0, 1, 0, 1, 0, 0), alpha = 0.05)
CIs_nonloc  <- CI(fit_melanoma, new_time, c(1, 1, 0, 1, 0, 0), alpha = 0.05)
```

```{r melanoma-hazard-plot}
ylim_h <- range(c(CIs_loc$lower_h, CIs_nonloc$upper_h), na.rm = TRUE)
plot(CIs_loc$time, CIs_loc$h, type = "l", col = "steelblue",
     ylim = ylim_h, xlab = "Time (months)", ylab = "Hazard",
     main = "Estimated hazard — melanoma, GH model")
lines(CIs_loc$time,    CIs_loc$lower_h,    col = "steelblue", lty = 2)
lines(CIs_loc$time,    CIs_loc$upper_h,    col = "steelblue", lty = 2)
lines(CIs_nonloc$time, CIs_nonloc$h,       col = "firebrick")
lines(CIs_nonloc$time, CIs_nonloc$lower_h, col = "firebrick", lty = 2)
lines(CIs_nonloc$time, CIs_nonloc$upper_h, col = "firebrick", lty = 2)
legend("topright", c("Localised", "Non-localised"),
       col = c("steelblue", "firebrick"), lty = 1)
```

```{r melanoma-survival-plot}
plot(CIs_loc$time, CIs_loc$S, type = "l", col = "steelblue",
     ylim = c(0, 1), xlab = "Time (months)", ylab = "Survival",
     main = "Estimated survival — melanoma, GH model")
lines(CIs_loc$time,    CIs_loc$lower_S,    col = "steelblue", lty = 2)
lines(CIs_loc$time,    CIs_loc$upper_S,    col = "steelblue", lty = 2)
lines(CIs_nonloc$time, CIs_nonloc$S,       col = "firebrick")
lines(CIs_nonloc$time, CIs_nonloc$lower_S, col = "firebrick", lty = 2)
lines(CIs_nonloc$time, CIs_nonloc$upper_S, col = "firebrick", lty = 2)
legend("topright", c("Localised", "Non-localised"),
       col = c("steelblue", "firebrick"), lty = 1)
```

### Time-varying hazard ratio

```{r melanoma-hr-plot}
hr <- CIs_nonloc$h / CIs_loc$h
plot(new_time, hr, type = "l", col = "purple",
     xlim = c(0, 200), ylim = c(0, max(hr[new_time <= 200], na.rm = TRUE)),
     xlab = "Time (months)", ylab = "Hazard ratio",
     main = "Time-varying HR — non-localised vs localised")
abline(h = 1, lty = 2, col = "grey50")
```

With only two parameters for the stage effect, the GH model captures the
time-varying hazard ratio: high at diagnosis, levelling off after approximately
6 years. This can be compared using survPen.

---

## Smoothing-parameter selection

The smoothing parameter $\lambda$ is chosen by minimising the modified LCV
criterion. Three strategies are available via `lcv_method`:

| `lcv_method` | Method | Notes |
|---|---|---|
| `"full"` (default) | Root-find on full LCV gradient (third-derivative correction) | Most accurate |
| `"approx"` | Root-find on first-order LCV gradient | Faster |
| `"optimize"` | Direct `optimize()` on LCV, no gradient | Gradient-free |

```{r lcv-methods, eval=FALSE}
# First-order gradient (faster)
fit <- fit_genhaz(..., profile = TRUE, lcv_method = "approx")

# Pure optimisation (no gradient)
fit <- fit_genhaz(..., profile = TRUE, lcv_method = "optimize")
```
